---
title: "Step 2: PacBio HiFi Quality Filtering"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
  github_document:
    toc: true
    toc_depth: 3
    html_preview: false
---

```{r setup, include=FALSE}
# Execute the analysis for the HTML report. The GitHub document remains a
# portable code/reference view and therefore does not rerun the workflow.
render_target <- knitr::opts_knit$get("rmarkdown.pandoc.to")
knitr::opts_chunk$set(
  eval = !(grepl("^gfm", render_target) || grepl("markdown_github", render_target, fixed = TRUE)),
  #eval = FALSE,
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

```{css custom-styles, echo=FALSE, eval=knitr::is_html_output()}
body {
  font-size: 16px;
  line-height: 1.6;
}

h1, h2, h3 {
  color: #2c3e50;
  margin-top: 1.5em;
}

code {
  background-color: #F2F2F2;
  color: #2c3e50;
  padding: 2px 6px;
  border-radius: 3px;
  font-size: 14px;
}

pre, pre code {
  font-size: 14px;
}

.alert-info {
  background-color: #2c3e50;
  color: #ffffff;
  border-left: 4px solid #f39c12;
  padding: 12px;
  margin: 15px 0;
}

.alert-warning {
  background-color: #2c3e50;
  color: #ffffff;
  border-left: 4px solid #e74c3c;
  padding: 12px;
  margin: 15px 0;
}

.alert-success {
  background-color: #d4edda;
  color: #155724;
  border-left: 4px solid #28a745;
  padding: 12px;
  margin: 15px 0;
}

table,
th,
td {
  text-align: left !important;
}

th, td {
  white-space: normal;
  word-wrap: break-word;
  overflow-wrap: anywhere;
  word-break: break-word;
  vertical-align: top;
}

/* Keep column names on one line; the DataTables viewport supplies horizontal
   scrolling when the resulting table is wider than the report body. */
th,
.dataTables_wrapper table.dataTable th {
  white-space: nowrap !important;
  word-break: normal !important;
  overflow-wrap: normal !important;
}

.dataTables_wrapper table.dataTable td,
.dataTables_wrapper .dt-left,
.dataTables_wrapper .dt-right,
.dataTables_wrapper .dt-center {
  text-align: left !important;
  white-space: normal !important;
  overflow-wrap: anywhere;
  word-break: break-word;
}

.dataTables_wrapper table.dataTable thead th {
  white-space: nowrap !important;
  word-break: normal !important;
  overflow-wrap: normal !important;
}

.dataTables_scrollBody,
.dataTables_scrollHead {
  overflow-x: auto !important;
}
```

# Introduction

## Purpose

This notebook is **Step 2** of the PacBio HiFi full-length 16S rRNA workflow. It reads the validated, sample-named FASTQ files generated by [Step 1](1_data_integrity_and_sample_mapping.md), implements PacBio-style read statistics and Q20 filtering in R, and creates an auditable QC report before [Step 3 (Primer Trimming with Cutadapt)](3_primer_trimming.md). After primer trimming, [Step 4](4_dada2_parameter_selection.md) supports `maxEE` selection and [Step 5](5_dada2_pipeline.md) performs the complete DADA2 analysis.

PacBio HiFi reads already span the full amplified 16S region. This notebook therefore assesses complete-read length, GC content, and quality value (QV) and does not choose a positional truncation length. Reads below the configurable Q20 threshold are excluded from the filtered FASTQs used by subsequent steps.

## Prerequisites

Before running this notebook:

- Complete [Step 1](1_data_integrity_and_sample_mapping.md) successfully.
- Confirm that [mapped_fastq/](../../results/1_data_integrity_and_sample_mapping/mapped_fastq/) contains the validated sample FASTQs.
- Confirm that [data_integrity_and_sample_mapping.xlsx](../../results/1_data_integrity_and_sample_mapping/data_integrity_and_sample_mapping.xlsx) contains the `Sample_Manifest` worksheet.
- Install the R dependencies used by the workflow.

## What This Notebook Does

1. **Reads the Step 1 manifest** and reconstructs portable paths to the validated FASTQ files.
2. **Streams every FASTQ** in fixed-size chunks.
3. **Calculates read statistics**, including read length, GC percentage, and quality value (QV) derived from mean base-error probability.
4. **Applies a quality threshold of Q20**, writing passing records to new filtered FASTQ files
5. **Adds the complete per-read audit table to Excel** and retains a bounded plotting sample so large runs remain memory efficient.
6. **Creates pre-filter read-length and QV distributions** as inline figures and standalone PDFs.
7. **Exports a documented Excel workbook** with run settings, pre/post-filter sample QC, and a Column Dictionary.

## Expected Input

- [results/1_data_integrity_and_sample_mapping/data_integrity_and_sample_mapping.xlsx](../../results/1_data_integrity_and_sample_mapping/data_integrity_and_sample_mapping.xlsx)
- [results/1_data_integrity_and_sample_mapping/mapped_fastq/](../../results/1_data_integrity_and_sample_mapping/mapped_fastq/)

## Expected Output

- [results/2_quality_filtered_reads/quality_and_read_length_summary.xlsx](../../results/2_quality_filtered_reads/quality_and_read_length_summary.xlsx)
- [results/2_quality_filtered_reads/Q20_filtered_fastq/](../../results/2_quality_filtered_reads/Q20_filtered_fastq/)
- [results/2_quality_filtered_reads/read_length_distribution.pdf](../../results/2_quality_filtered_reads/read_length_distribution.pdf)
- [results/2_quality_filtered_reads/read_quality_distribution.pdf](../../results/2_quality_filtered_reads/read_quality_distribution.pdf)

------------------------------------------------------------------------

# Environment Setup

## Load Required Packages {#load-packages}

Load the packages used for streaming FASTQ files, building plots and tables, reading the [Step 1](1_data_integrity_and_sample_mapping.md) manifest, and writing documented Excel results. The shared helper functions are sourced from [R/functions/](../functions/).

```{r load-packages, message=FALSE, warning=FALSE}
# Declare dependencies in one place so the notebook fails before reading or
# writing files when the environment is incomplete.
required_packages <- c(
  "data.table", "DT", "fs", "ggplot2", "here", "knitr",
  "openxlsx", "readxl", "ShortRead"
)

missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, quietly = TRUE, FUN.VALUE = logical(1))
]

if (length(missing_packages) > 0L) {
  stop(
    "Missing required package(s): ", paste(missing_packages, collapse = ", "),
    ". Install them before running Notebook 2."
  )
}

# Attach packages used repeatedly; less frequent calls remain namespace-qualified.
library(data.table)
library(fs)
library(ggplot2)
library(here)

# Reuse the same workbook and portable-link behavior as Notebook 1.
source(here("R", "functions", "add_sheet_to_excel_function.R"))
source(here("R", "functions", "build_column_dictionary_function.R"))
source(here("R", "functions", "render_output_links_function.R"))
source(here("R", "functions", "render_output_tree_function.R"))
```

------------------------------------------------------------------------

# Configuration

## Define Input and Output Paths {#define-paths}

Define all paths relative to the project root. The [Step 1](1_data_integrity_and_sample_mapping.md) workbook supplies sample identities, while the FASTQ basenames are resolved inside the current project so copied projects remain portable.

```{r define-paths}
# Step 1 inputs used by this notebook.
step1_folder <- here("results", "1_data_integrity_and_sample_mapping")
step1_workbook <- here(step1_folder, "data_integrity_and_sample_mapping.xlsx")
mapped_fastq_folder <- here(step1_folder, "mapped_fastq")

# Step 2 owns this complete output subtree.
results_folder <- here("results", "2_quality_filtered_reads")
filtered_fastq_folder <- here(results_folder, "Q20_filtered_fastq")
# The per-read table is streamed through a temporary TSV to keep memory bounded,
# imported into Excel, and deleted before the notebook finishes.
read_statistics_path <- tempfile(pattern = "step2_per_read_", fileext = ".tsv")
workbook_path <- here(results_folder, "quality_and_read_length_summary.xlsx")
length_plot_path <- here(results_folder, "read_length_distribution.pdf")
quality_plot_path <- here(results_folder, "read_quality_distribution.pdf")
```

## Define QC Parameters {#define-qc-parameters}

Set transparent review thresholds and streaming limits. The expected length range should match the primer pair and biological target; it is used only for visualization and does not filter reads!

```{r define-qc-parameters}
# Full-length bacterial 16S amplicons are commonly near 1.5 kb. These broad
# review bounds can be adjusted for another primer pair or target region.
expected_length_min <- 1200L
expected_length_max <- 1800L

# Match the user-adjustable QV filter in PacBio's HiFi-16S workflow. QV is
# calculated from the mean base-error probability, as documented by SeqKit.
read_qv_cutoff <- 20

# Read FASTQ records in bounded chunks and keep at most this many observations
# per sample for plots and approximate distribution quantiles.
stream_yield_size <- 10000L
maximum_plot_reads_per_sample <- 50000L

# Phred+33 is the standard FASTQ quality encoding used for these files.
quality_ascii_offset <- 33L
```

## Validate Configuration {#validate-configuration}

Validate inputs and parameter relationships before creating or replacing any Step 2 output.

```{r validate-configuration}
# Fail early when Step 1 artifacts are missing.
if (!file_exists(step1_workbook)) {
  stop("Step 1 workbook does not exist: ", step1_workbook)
}
if (!dir_exists(mapped_fastq_folder)) {
  stop("Step 1 mapped FASTQ folder does not exist: ", mapped_fastq_folder)
}

# Prevent invalid thresholds or non-positive streaming limits.
if (expected_length_min < 1L || expected_length_max <= expected_length_min) {
  stop("Expected read-length bounds must satisfy 1 <= minimum < maximum.")
}
if (stream_yield_size < 1L || maximum_plot_reads_per_sample < 1L) {
  stop("Streaming and plotting limits must be positive integers.")
}

# Create only this notebook's managed output folders.
dir_create(results_folder, recurse = TRUE)
dir_create(filtered_fastq_folder, recurse = TRUE)
```

The resolved inputs, managed result folder, quality threshold, and streaming settings are printed below so this report records the exact run configuration.

```{r report-configuration, echo=FALSE}
cat("Step 1 workbook:", step1_workbook, "\n")
cat("Mapped FASTQ folder:", mapped_fastq_folder, "\n")
cat("Step 2 results folder:", results_folder, "\n")
cat("Expected read-length range:", expected_length_min, "to", expected_length_max, "bp\n")
cat("Per-read QV filtering cutoff: Q", read_qv_cutoff, "\n", sep = "")
```

------------------------------------------------------------------------

# Functions

## Define Table and FASTQ Helpers {#helper-functions}

Define the shared presentation behavior and a streaming FASTQ summarizer. Comments describe the memory bounds, quality conversion, and distinction between full-dataset summaries and bounded plotting samples.

```{r helper-functions}
render_table <- function(x, caption = NULL, column_widths = NULL) {
  # Use one normalized data.frame for both HTML and GitHub renderers.
  table_data <- as.data.frame(x)

  if (knitr::is_html_output()) {
    # Force every column left and optionally widen selected columns. DataTables
    # uses zero-based targets, so subtract one from R column positions.
    column_definitions <- list(list(className = "dt-left", targets = "_all"))
    if (!is.null(column_widths)) {
      width_columns <- which(!is.na(column_widths) & nzchar(column_widths))
      column_definitions <- c(
        column_definitions,
        lapply(width_columns, function(column_index) {
          list(targets = column_index - 1L, width = column_widths[[column_index]])
        })
      )
    }

    # These are inspection tables: retain scrolling but remove sorting, search,
    # pagination, and entry-count controls.
    DT::datatable(
      table_data,
      caption = caption,
      rownames = FALSE,
      class = "display compact",
      options = list(
        scrollX = TRUE,
        scrollY = "400px",
        scrollCollapse = TRUE,
        paging = FALSE,
        searching = FALSE,
        ordering = FALSE,
        info = FALSE,
        dom = "t",
        autoWidth = !is.null(column_widths),
        columnDefs = column_definitions
      )
    )
  } else {
    # GitHub receives a portable, fully left-aligned Markdown table.
    knitr::kable(
      table_data,
      caption = caption,
      row.names = FALSE,
      align = rep("l", ncol(table_data))
    )
  }
}

calculate_read_qv <- function(quality_strings) {
  # SeqKit defines average read quality by converting every base Q score to an
  # error probability, averaging those probabilities, and converting the mean
  # error probability back to Phred scale. This is deliberately not the
  # arithmetic mean of the base Q scores.
  vapply(
    quality_strings,
    function(quality_string) {
      base_q <- utf8ToInt(quality_string) - quality_ascii_offset
      -10 * log10(mean(10^(-base_q / 10)))
    },
    numeric(1)
  )
}

summarize_fastq <- function(path, sample_id, filtered_path, statistics_connection) {
  # Stream a fixed number of records per yield so memory use is independent of
  # total FASTQ size. The streamer is always closed, including after an error.
  streamer <- ShortRead::FastqStreamer(path, n = stream_yield_size)
  on.exit(close(streamer), add = TRUE)

  total_reads <- 0L
  total_bases <- 0
  total_length_squared <- 0
  minimum_length <- Inf
  maximum_length <- -Inf
  reads_in_expected_range <- 0L
  total_read_qv <- 0
  total_read_qv_squared <- 0
  minimum_read_qv <- Inf
  maximum_read_qv <- -Inf
  total_gc_percent <- 0
  reads_with_defined_gc <- 0L
  passing_reads <- 0L
  passing_bases <- 0

  # Plot vectors are intentionally capped. Full-dataset means, ranges, counts,
  # and standard deviations continue to use every streamed read.
  plot_lengths <- numeric(0)
  plot_read_qv <- numeric(0)

  # Remove a prior filtered file before the first append. Step 1 inputs remain
  # read-only and can therefore always be used to reproduce the filtering.
  if (file_exists(filtered_path)) file_delete(filtered_path)

  repeat {
    fastq_chunk <- ShortRead::yield(streamer)
    if (length(fastq_chunk) == 0L) break

    sequences <- ShortRead::sread(fastq_chunk)
    read_lengths <- as.numeric(Biostrings::width(sequences))
    # ShortRead stores qualities as a FastqQuality wrapper around a BStringSet.
    # Access the documented S4 slots directly because quality() is a generic
    # attached by ShortRead but is not exported for namespace-qualified use.
    quality_strings <- as.character(slot(slot(fastq_chunk, "quality"), "quality"))
    read_qv <- calculate_read_qv(quality_strings)

    # SeqKit reports GC as (G+C)/(A+C+G+T), excluding ambiguous characters
    # from the denominator. Reads without canonical bases receive NA.
    base_counts <- Biostrings::letterFrequency(sequences, letters = c("A", "C", "G", "T"))
    canonical_bases <- rowSums(base_counts)
    gc_percent <- fifelse(
      canonical_bases > 0,
      100 * (base_counts[, "C"] + base_counts[, "G"]) / canonical_bases,
      NA_real_
    )
    pass_qv <- read_qv >= read_qv_cutoff

    chunk_reads <- length(read_lengths)
    total_reads <- total_reads + chunk_reads
    total_bases <- total_bases + sum(read_lengths)
    total_length_squared <- total_length_squared + sum(read_lengths^2)
    minimum_length <- min(minimum_length, min(read_lengths))
    maximum_length <- max(maximum_length, max(read_lengths))
    reads_in_expected_range <- reads_in_expected_range + sum(
      read_lengths >= expected_length_min & read_lengths <= expected_length_max
    )

    total_read_qv <- total_read_qv + sum(read_qv)
    total_read_qv_squared <- total_read_qv_squared + sum(read_qv^2)
    minimum_read_qv <- min(minimum_read_qv, min(read_qv))
    maximum_read_qv <- max(maximum_read_qv, max(read_qv))
    total_gc_percent <- total_gc_percent + sum(gc_percent, na.rm = TRUE)
    reads_with_defined_gc <- reads_with_defined_gc + sum(!is.na(gc_percent))
    passing_reads <- passing_reads + sum(pass_qv)
    passing_bases <- passing_bases + sum(read_lengths[pass_qv])

    # Write every per-read statistic to the uncompressed audit table. The table
    # mirrors the SeqKit fx2tab fields used by PacBio while remaining R-native.
    read_statistics <- data.table(
      SampleID = sample_id,
      ReadID = as.character(ShortRead::id(fastq_chunk)),
      ReadLength = read_lengths,
      GCPercent = round(gc_percent, 4),
      ReadQV = round(read_qv, 4),
      PassQVFilter = pass_qv
    )
    writeLines(
      apply(read_statistics, 1L, paste, collapse = "\t"),
      con = statistics_connection
    )

    # Append only QV-passing records to a new uncompressed FASTQ. An empty output is
    # still created below when no read passes, preserving one output per input.
    if (any(pass_qv)) {
      ShortRead::writeFastq(
        fastq_chunk[pass_qv],
        filtered_path,
        mode = "a",
        compress = FALSE
      )
    }

    # Retain the earliest bounded observations deterministically. Repeated runs
    # therefore generate identical plots from identical input files.
    remaining_plot_slots <- maximum_plot_reads_per_sample - length(plot_lengths)
    if (remaining_plot_slots > 0L) {
      keep <- seq_len(min(chunk_reads, remaining_plot_slots))
      plot_lengths <- c(plot_lengths, read_lengths[keep])
      plot_read_qv <- c(plot_read_qv, read_qv[keep])
    }
  }

  if (total_reads == 0L) stop("FASTQ contains no reads: ", path)

  # Preserve a one-output-per-input contract even when no read reaches Q20.
  if (!file_exists(filtered_path)) {
    empty_filtered_connection <- file(filtered_path, open = "wb")
    close(empty_filtered_connection)
  }

  # Compute full-dataset means and sample standard deviations from accumulated
  # sums, avoiding storage of every observation.
  mean_length <- total_bases / total_reads
  length_variance <- if (total_reads > 1L) {
    (total_length_squared - total_reads * mean_length^2) / (total_reads - 1L)
  } else {
    0
  }
  mean_read_qv <- total_read_qv / total_reads
  quality_variance <- if (total_reads > 1L) {
    (total_read_qv_squared - total_reads * mean_read_qv^2) / (total_reads - 1L)
  } else {
    0
  }

  # Quantiles are calculated from the documented bounded plotting sample. This
  # is exact whenever total_reads does not exceed maximum_plot_reads_per_sample.
  length_quantiles <- stats::quantile(
    plot_lengths,
    probs = c(0.05, 0.25, 0.50, 0.75, 0.95),
    names = FALSE,
    type = 7
  )

  list(
    summary = data.table(
      SampleID = sample_id,
      FASTQ = as.character(path),
      TotalReads = as.numeric(total_reads),
      TotalBases = as.numeric(total_bases),
      MinimumLength = as.numeric(minimum_length),
      MeanLength = round(mean_length, 1),
      LengthSD = round(sqrt(max(length_variance, 0)), 1),
      MaximumLength = as.numeric(maximum_length),
      LengthP05 = round(length_quantiles[[1]], 1),
      LengthP25 = round(length_quantiles[[2]], 1),
      MedianLength = round(length_quantiles[[3]], 1),
      LengthP75 = round(length_quantiles[[4]], 1),
      LengthP95 = round(length_quantiles[[5]], 1),
      ReadsInExpectedLengthRange = as.numeric(reads_in_expected_range),
      PercentInExpectedLengthRange = round(100 * reads_in_expected_range / total_reads, 2),
      MeanGCPercent = if (reads_with_defined_gc > 0L) {
        round(total_gc_percent / reads_with_defined_gc, 2)
      } else {
        NA_real_
      },
      MinimumReadQV = round(minimum_read_qv, 2),
      MeanReadQV = round(mean_read_qv, 2),
      ReadQVSD = round(sqrt(max(quality_variance, 0)), 2),
      MaximumReadQV = round(maximum_read_qv, 2),
      QVCutoff = as.numeric(read_qv_cutoff),
      PassingReads = as.numeric(passing_reads),
      FilteredReads = as.numeric(total_reads - passing_reads),
      PercentPassingQV = round(100 * passing_reads / total_reads, 2),
      PassingBases = as.numeric(passing_bases),
      FilteredFASTQ = as.character(filtered_path),
      PlotReads = as.numeric(length(plot_lengths))
    ),
    plot_data = data.table(
      SampleID = sample_id,
      ReadLength = plot_lengths,
      ReadQV = plot_read_qv
    )
  )
}
```

------------------------------------------------------------------------

# Step 1 Manifest

## Read and Validate the Sample Manifest {#read-manifest}

Read sample identities from the [Step 1 audit workbook](../../results/1_data_integrity_and_sample_mapping/data_integrity_and_sample_mapping.xlsx) rather than inferring them again from filenames. Reconstruct each FASTQ path inside the current project and validate the one-row-per-sample relationship.

```{r read-manifest}
# Read only the fields required for QC and path reconstruction.
manifest_raw <- readxl::read_excel(
  step1_workbook,
  sheet = "Sample_Manifest",
  col_types = "text"
)
manifest_raw <- as.data.table(manifest_raw)

required_manifest_columns <- c("PoolNumber", "SampleID", "MappedFASTQ", "IntegrityStatus")
missing_manifest_columns <- setdiff(required_manifest_columns, names(manifest_raw))
if (length(missing_manifest_columns) > 0L) {
  stop("Step 1 Sample_Manifest is missing: ", paste(missing_manifest_columns, collapse = ", "))
}

# Refuse QC input from a Step 1 row that did not pass integrity validation.
if (any(manifest_raw$IntegrityStatus != "PASS")) {
  stop("Every Step 1 Sample_Manifest row must have IntegrityStatus = PASS.")
}

sample_manifest <- manifest_raw[, .(
  PoolNumber = as.integer(PoolNumber),
  SampleID = as.character(SampleID),
  FASTQ = as.character(path(mapped_fastq_folder, path_file(MappedFASTQ)))
)]
setorder(sample_manifest, PoolNumber, SampleID)

if (nrow(sample_manifest) == 0L) stop("Step 1 Sample_Manifest contains no samples.")
if (anyDuplicated(sample_manifest$SampleID)) stop("Step 1 Sample_Manifest contains duplicate SampleIDs.")
if (any(!file_exists(sample_manifest$FASTQ))) {
  stop(
    "Mapped FASTQ file(s) are missing: ",
    paste(sample_manifest[!file_exists(FASTQ), FASTQ], collapse = ", ")
  )
}

render_table(
  sample_manifest,
  "Validated Step 1 inputs for PacBio QC",
  column_widths = c("120px", "190px", "520px")
)
```

This status line confirms how many Step 1 samples have valid, uniquely resolved FASTQ inputs for quality assessment.

```{r report-input-count, echo=FALSE}
cat("Validated FASTQ inputs:", nrow(sample_manifest), "\n")
```

------------------------------------------------------------------------

# FASTQ Quality Assessment

## Stream and Summarize FASTQ Files {#summarize-fastq}

Stream every validated FASTQ, calculate full-dataset statistics, and write reads meeting the configured QV threshold to new FASTQ files.

```{r summarize-fastq-files}
# Rebuild the uncompressed per-read audit table on every run and write the header
# once. A single open connection permits memory-bounded chunk-by-chunk output.
if (file_exists(read_statistics_path)) file_delete(read_statistics_path)

# This notebook owns Q20_filtered_fastq/. Remove FASTQs from an earlier sample
# naming scheme before writing the current dataset so obsolete identifiers
# cannot coexist with the newly generated workflow inputs.
stale_filtered_fastq <- dir_ls(
  filtered_fastq_folder,
  regexp = "\\.(fastq|fq)$",
  type = "file",
  recurse = FALSE
)
if (length(stale_filtered_fastq) > 0L) file_delete(stale_filtered_fastq)

statistics_connection <- file(read_statistics_path, open = "wt")
writeLines(
  paste(c("SampleID", "ReadID", "ReadLength", "GCPercent", "ReadQV", "PassQVFilter"), collapse = "\t"),
  statistics_connection
)

# Keep the connection safe on errors and close it normally after all samples.
qc_results <- tryCatch(
  lapply(seq_len(nrow(sample_manifest)), function(i) {
    filtered_path <- path(
      filtered_fastq_folder,
      paste0(sample_manifest$SampleID[[i]], "_filtered_Q", read_qv_cutoff, ".fastq")
    )
    summarize_fastq(
      path = sample_manifest$FASTQ[[i]],
      sample_id = sample_manifest$SampleID[[i]],
      filtered_path = filtered_path,
      statistics_connection = statistics_connection
    )
  }),
  finally = close(statistics_connection)
)

sample_qc <- rbindlist(lapply(qc_results, `[[`, "summary"), use.names = TRUE)
plot_data <- rbindlist(lapply(qc_results, `[[`, "plot_data"), use.names = TRUE)

# Restore pool metadata and deterministic sample order after streaming.
sample_qc <- merge(
  sample_manifest[, .(PoolNumber, SampleID)],
  sample_qc,
  by = "SampleID",
  all.x = TRUE,
  sort = FALSE
)
setorder(sample_qc, PoolNumber, SampleID)
plot_data[, SampleID := factor(SampleID, levels = sample_qc$SampleID)]

# Confirm that each sample retained reads at Q20 and assess the broad expected
# length interval independently. Filtering is per read, not based on a sample's
# mean QV.
sample_qc[, QualityStatus := fifelse(
  PassingReads > 0,
  "PASS",
  "REVIEW: no reads passed the configured QV cutoff"
)]
sample_qc[, LengthStatus := fifelse(
  PercentInExpectedLengthRange >= 50,
  "PASS",
  "REVIEW: fewer than 50% of reads inside expected range"
)]
sample_qc[, OverallQCStatus := fifelse(
  QualityStatus == "PASS" & LengthStatus == "PASS",
  "PASS",
  "REVIEW"
)]
```

The report now confirms how many FASTQs were processed and how many reproducibly sampled observations are available for the distribution plots.

```{r report-qc-completion, echo=FALSE}
cat("FASTQ files summarized:", nrow(sample_qc), "\n")
cat("Reads retained for plotting:", nrow(plot_data), "\n")
```

## Sample QC Summary {#sample-qc-summary}

Display the primary per-sample QC metrics.

```{r display-qc-summary}
render_table(
  sample_qc[, .(
    PoolNumber,
    SampleID,
    TotalReads,
    MeanLength,
    MedianLength,
    MeanGCPercent,
    PercentInExpectedLengthRange,
    MeanReadQV,
    PassingReads,
    FilteredReads,
    PercentPassingQV,
    QualityStatus,
    LengthStatus,
    OverallQCStatus
  )],
  "PacBio HiFi sample quality summary",
  column_widths = c("80px", "140px", rep(NA_character_, 12))
)
```

## Read-Length Distribution {#length-distribution}

Plot the bounded read-length sample for each biological sample. Dashed vertical lines mark the configured expected range; they are merely visualization guides, not filtering boundaries!

```{r plot-read-lengths}
length_plot <- ggplot(plot_data, aes(x = ReadLength, fill = SampleID)) +
  geom_histogram(bins = 60, alpha = 0.75, color = "white") +
  geom_vline(
    xintercept = c(expected_length_min, expected_length_max),
    linetype = "dashed",
    color = "#C0392B",
    linewidth = 0.6
  ) +
  geom_text(
    data = sample_qc,
    aes(
      x = Inf,
      y = Inf,
      label = paste0(
        "Total reads: ", TotalReads,
        "\nInside range: ", ReadsInExpectedLengthRange
      )
    ),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.25,
    size = 3.4
  ) +
  facet_wrap(~ SampleID, scales = "free_y", ncol = 2) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(
    title = "PacBio HiFi read-length distributions",
    subtitle = paste0(
      "Dashed lines show the configured review range: ",
      expected_length_min, "–", expected_length_max, " bp"
    ),
    x = "Read length (bp)",
    y = "Number of reads"
  ) +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold"))

print(length_plot)
```

The read-length distribution displayed above is also saved as a standalone PDF, and its path is reported below.

```{r save-read-length-plot, echo=FALSE}
ggsave(length_plot_path, length_plot, width = 11, height = 8.5, units = "in")
cat("Read-length plot saved to:", length_plot_path, "\n")
```

## Read-QV Distribution {#quality-distribution}

Plot the read-quality distribution calculated from each read's mean base-error probability. The dashed line is the active Q20 filtering threshold!

```{r plot-read-quality}
quality_plot <- ggplot(plot_data, aes(x = ReadQV, fill = SampleID)) +
  geom_histogram(bins = 60, alpha = 0.75, color = "white") +
  geom_vline(
    xintercept = read_qv_cutoff,
    linetype = "dashed",
    color = "#C0392B",
    linewidth = 0.6
  ) +
  geom_text(
    data = sample_qc,
    aes(
      x = Inf,
      y = Inf,
      label = paste0(
        "Total reads: ", TotalReads,
        "\nReads with Q ≥ ", QVCutoff, ": ", PassingReads
      )
    ),
    inherit.aes = FALSE,
    hjust = 1.05,
    vjust = 1.25,
    size = 3.4
  ) +
  facet_wrap(~ SampleID, scales = "free_y", ncol = 2) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(
    title = "PacBio HiFi per-read QV distributions",
    subtitle = paste0(
      "Dashed line shows the active filtering cutoff: Q",
      read_qv_cutoff
    ),
    x = "Read quality",
    y = "Number of reads"
  ) +
  theme_minimal(base_size = 12) +
  theme(strip.text = element_text(face = "bold"))

print(quality_plot)
```

The read-quality distribution displayed above is saved as a standalone PDF for sharing and comparison outside the notebook.

```{r save-quality-plot, echo=FALSE}
ggsave(quality_plot_path, quality_plot, width = 11, height = 8.5, units = "in")
cat("Read-QV plot saved to:", quality_plot_path, "\n")
```

------------------------------------------------------------------------

# Export Results

## Save QC Tables to Excel {#save-excel}

Create a typed run summary and a complete per-sample QC table, write both with the shared [add_sheet_to_excel_function.R](../functions/add_sheet_to_excel_function.R) helper, and document every exported column with [build_column_dictionary_function.R](../functions/build_column_dictionary_function.R).

```{r export-results, echo=FALSE}
# Record one row per sample and repeat the configured parameters so each row is
# independently interpretable when the worksheet is filtered or copied.
run_summary <- sample_qc[, .(
  PoolNumber = as.integer(PoolNumber),
  SampleID = as.character(SampleID),
  TotalReads = as.numeric(TotalReads),
  TotalBases = as.numeric(TotalBases),
  PassingReads = as.numeric(PassingReads),
  FilteredReads = as.numeric(FilteredReads),
  PercentPassingQV = as.numeric(PercentPassingQV),
  ExpectedLengthMinimum = as.integer(expected_length_min),
  ExpectedLengthMaximum = as.integer(expected_length_max),
  ReadQVCutoff = as.numeric(read_qv_cutoff),
  StreamYieldSize = as.integer(stream_yield_size),
  MaximumPlotReadsPerSample = as.integer(maximum_plot_reads_per_sample),
  OverallQCStatus = as.character(OverallQCStatus)
)]
setorder(run_summary, PoolNumber, SampleID)

# The Sample_QC worksheet begins with pool and sample identifiers. Runtime file
# paths are intentionally omitted because they are machine-specific and are
# already represented by the project-relative output structure.
sample_qc_export <- copy(sample_qc)
sample_qc_export[, c("FASTQ", "FilteredFASTQ") := NULL]
setcolorder(
  sample_qc_export,
  c("PoolNumber", "SampleID", setdiff(names(sample_qc_export), c("PoolNumber", "SampleID")))
)

# Include the complete per-read audit table in the workbook so no standalone
# tabular result is required. Excel worksheets can contain at most 1,048,576
# rows including the header.
per_read_statistics <- fread(read_statistics_path)
if (nrow(per_read_statistics) > 1048575L) {
  stop(
    "Per_Read_Statistics exceeds the Excel worksheet row limit. ",
    "Split the analysis into smaller runs before exporting."
  )
}

column_definitions <- c(
  ExpectedLengthMinimum = "Lower bound of the configured read-length review interval in base pairs.",
  ExpectedLengthMaximum = "Upper bound of the configured read-length review interval in base pairs.",
  ReadQVCutoff = "Minimum per-read QV retained in filtered FASTQ output; default Q20 follows PacBio's HiFi-16S workflow.",
  StreamYieldSize = "Maximum FASTQ records read into memory per streaming yield.",
  MaximumPlotReadsPerSample = "Maximum deterministic observations retained per sample for plots and quantiles.",
  PoolNumber = "Numeric PacBio pool number inherited from the Step 1 Sample_Manifest.",
  SampleID = "Stable sample identifier inherited from the Step 1 Sample_Manifest.",
  ReadID = "FASTQ record identifier.",
  ReadLength = "Read length in base pairs.",
  GCPercent = "Per-read GC percentage calculated as (G+C)/(A+C+G+T).",
  ReadQV = "Per-read QV calculated from mean base-error probability using the SeqKit method.",
  PassQVFilter = "TRUE when ReadQV is at or above the configured QV cutoff.",
  TotalReads = "Total FASTQ records in the sample.",
  TotalBases = "Total bases across all reads in the sample FASTQ.",
  MinimumLength = "Shortest read length in base pairs.",
  MeanLength = "Arithmetic mean read length in base pairs across all reads.",
  LengthSD = "Sample standard deviation of read length across all reads.",
  MaximumLength = "Longest read length in base pairs.",
  LengthP05 = "Fifth percentile read length estimated from the bounded plotting sample.",
  LengthP25 = "Twenty-fifth percentile read length estimated from the bounded plotting sample.",
  MedianLength = "Median read length estimated from the bounded plotting sample.",
  LengthP75 = "Seventy-fifth percentile read length estimated from the bounded plotting sample.",
  LengthP95 = "Ninety-fifth percentile read length estimated from the bounded plotting sample.",
  ReadsInExpectedLengthRange = "Reads between the configured expected minimum and maximum lengths, inclusive.",
  PercentInExpectedLengthRange = "Percentage of all sample reads inside the configured expected length interval.",
  MeanGCPercent = "Arithmetic mean of per-read GC percentage, calculated as (G+C)/(A+C+G+T).",
  MinimumReadQV = "Lowest SeqKit-equivalent per-read QV observed in the sample.",
  MeanReadQV = "Arithmetic mean of the SeqKit-equivalent per-read QVs in the sample.",
  ReadQVSD = "Sample standard deviation of SeqKit-equivalent per-read QV.",
  MaximumReadQV = "Highest SeqKit-equivalent per-read QV observed in the sample.",
  QVCutoff = "Minimum per-read QV retained in the filtered FASTQ for this sample.",
  PassingReads = "Input reads with ReadQV at or above QVCutoff.",
  FilteredReads = "Input reads with ReadQV below QVCutoff.",
  PercentPassingQV = "Percentage of input reads retained by the QV filter.",
  PassingBases = "Total bases among reads retained by the QV filter.",
  PlotReads = "Reads retained in the bounded deterministic plotting and quantile sample.",
  QualityStatus = "PASS when at least one read passes the configured QV filter; otherwise REVIEW.",
  LengthStatus = "PASS when at least half of reads fall inside the expected length range; otherwise REVIEW.",
  OverallQCStatus = "PASS when both QualityStatus and LengthStatus pass; otherwise REVIEW."
)

# Rebuild the workbook from the current run so stale worksheets cannot survive.
if (file_exists(workbook_path)) file_delete(workbook_path)

workbook_sheets <- list(
  Run_Summary = run_summary,
  Sample_QC = sample_qc_export,
  Per_Read_Statistics = per_read_statistics
)

for (sheet_name in names(workbook_sheets)) {
  add_sheet_to_excel(
    workbook_path = workbook_path,
    sheet_name = sheet_name,
    data = workbook_sheets[[sheet_name]],
    rownames = FALSE,
    overwrite = FALSE
  )
}

# Build the dictionary and write matching header comments on both source sheets.
column_dictionary <- do.call(
  rbind,
  lapply(names(workbook_sheets), function(sheet_name) {
    build_column_dictionary(
      sheet_name = sheet_name,
      data = workbook_sheets[[sheet_name]],
      descriptions = column_definitions,
      workbook_path = workbook_path,
      rownames = FALSE
    )
  })
)

dictionary_definitions <- c(
  Sheet = "Worksheet containing the documented column.",
  Column = "Column name exactly as written in the documented worksheet.",
  Explanation = "Definition and derivation of the documented column."
)

add_sheet_to_excel(
  workbook_path = workbook_path,
  sheet_name = "Column_Dictionary",
  data = column_dictionary,
  rownames = FALSE,
  overwrite = FALSE
)

# Suppress the returned dictionary table so it is written to Excel but is not
# displayed as an unintended table in the HTML report.
invisible(build_column_dictionary(
  sheet_name = "Column_Dictionary",
  data = column_dictionary,
  descriptions = dictionary_definitions,
  workbook_path = workbook_path,
  rownames = FALSE
))

# The temporary TSV has now been transferred into the workbook and is removed
# so the results folder contains no standalone tabular exports.
if (file_exists(read_statistics_path)) file_delete(read_statistics_path)
```

The report prints the completed workbook path after all result sheets, column definitions, and header comments have been written.

```{r report-export-path, echo=FALSE}
cat("QC workbook written to:", workbook_path, "\n")
```

## Final QC Decision {#final-decision}

Confirm that every input produced a complete summary and every required output file exists. A `REVIEW` status is not a processing failure because the thresholds are descriptive review rules rather than filters.

```{r final-qc-decision}
# Structural completion is required even when one or more biological samples
# are flagged for review.
qc_complete <-
  nrow(sample_qc) == nrow(sample_manifest) &&
  all(sample_qc$TotalReads > 0) &&
  all(file_exists(sample_qc$FilteredFASTQ)) &&
  file_exists(workbook_path) &&
  file_exists(length_plot_path) &&
  file_exists(quality_plot_path)

if (!qc_complete) {
  stop("Step 2 QC outputs are incomplete. Review the preceding diagnostics.")
}
```

After structural validation succeeds, this final status summarizes the number of samples, reads passing Q ≥ 20, and samples flagged for length-distribution review.

```{r report-final-status, echo=FALSE}
cat(
  "Step 2 complete —", nrow(sample_qc), "samples summarized;",
  sum(sample_qc$PassingReads), "reads retained at Q", read_qv_cutoff, ";",
  sum(sample_qc$OverallQCStatus == "PASS"), "passed configured review rules and",
  sum(sample_qc$OverallQCStatus != "PASS"), "require review.\n"
)
```

------------------------------------------------------------------------

# Output Files

The links below open the generated workbook and standalone PDF plots.

```{r output-links, echo=FALSE, results='asis'}
render_output_links(
  c(workbook_path, length_plot_path, quality_plot_path),
  labels = c(
    "PacBio quality and read-length summary workbook",
    "Read-length distribution plot (PDF)",
    "Read-QV distribution plot (PDF)"
  )
)
```

# Output File Summary

The live tree below lists the actual contents of [results/2_quality_filtered_reads/](../../results/2_quality_filtered_reads/) so the report cannot drift from files on disk.

```{r output-summary, echo=FALSE, results='asis'}
render_output_tree(
  results_folder,
  descriptions = c(
    "quality_and_read_length_summary.xlsx" = "Run settings, per-sample PacBio QC metrics, and Column Dictionary",
    "Q20_filtered_fastq" = "Uncompressed sample FASTQs containing reads retained at Q20",
    "read_length_distribution.pdf" = "Standalone read-length distribution plot",
    "read_quality_distribution.pdf" = "Standalone read-quality distribution plot"
  )
)
```

# Recommended Next Step

Review samples marked `REVIEW`, confirm that the expected length interval and QV cutoff are appropriate, and use the Q20-filtered FASTQs for the planned primer-removal notebook. The original mapped FASTQs remain unchanged. Proceed to [Step 3 (Primer Trimming with Cutadapt)](3_primer_trimming.md).

# Session Information

Record the exact R, platform, and package environment used to produce the report.

```{r session-info, echo=FALSE}
sessionInfo()
```

# References

## Methods and Software

- Callahan BJ et al. (2019). High-throughput amplicon sequencing of the full-length 16S rRNA gene with single-nucleotide resolution. *Nucleic Acids Research*, 47(18):e103. [Full-length 16S method](https://doi.org/10.1093/nar/gkz569)
- Morgan M et al. (2009). ShortRead: a Bioconductor package for input, quality assessment and exploration of high-throughput sequence data. *Bioinformatics*, 25(19):2607–2608. [ShortRead publication](https://doi.org/10.1093/bioinformatics/btp450)
- Wickham H (2016). *ggplot2: Elegant Graphics for Data Analysis*. Springer-Verlag New York. [ggplot2 documentation](https://ggplot2.tidyverse.org/)

## Documentation

- [Bioconductor ShortRead package](https://bioconductor.org/packages/release/bioc/html/ShortRead.html)
- [DADA2 full-length PacBio example](https://benjjneb.github.io/LRASManuscript/LRASms_fecal.html)
- [PacBio HiFi full-length 16S workflow](https://github.com/PacificBiosciences/HiFi-16S-workflow)
- [PacBio HiFi-16S pipeline overview](https://github.com/PacificBiosciences/HiFi-16S-workflow/blob/main/pipeline_overview.md)
- [SeqKit quality-statistics documentation](https://bioinf.shenwei.me/seqkit/usage/)
- [DT documentation](https://rstudio.github.io/DT/)
- [here package documentation](https://here.r-lib.org/)

## Related Project Files

- [Notebook 1 source](1_data_integrity_and_sample_mapping.Rmd)
- [Step 1 mapped FASTQs](../../results/1_data_integrity_and_sample_mapping/mapped_fastq/)
- [Step 1 audit workbook](../../results/1_data_integrity_and_sample_mapping/data_integrity_and_sample_mapping.xlsx)
- [Shared R functions](../functions/)

# Appendix: Troubleshooting Guide

## Step 1 Workbook or FASTQs Are Missing

Run [Step 1](1_data_integrity_and_sample_mapping.md) first and require its final integrity decision to pass. Do not reconstruct sample identities manually in Step 2.

## Mean Quality Values Look Unexpected

This notebook interprets FASTQ quality characters as Phred+33. It converts each base Q score to error probability, averages those probabilities within a read, and converts the result back to QV, matching SeqKit's documented method. Confirm the encoding of externally supplied FASTQs before changing `quality_ascii_offset`.

## Read Lengths Fall Outside the Expected Range

Confirm that `expected_length_min` and `expected_length_max` match the primer pair and target. A `REVIEW` flag does not remove reads; it indicates that primer removal and downstream filtering choices deserve inspection.

## Large Runs Use Too Much Plotting Memory

Reduce `maximum_plot_reads_per_sample`. Full-dataset counts, means, ranges, and standard deviations will still use every streamed read; only plotted observations and reported length quantiles use the bounded sample.
